---
title: The Man in the BoxPlot
author: Michelle Evans
date: '2018-04-04'
slug: the-man-in-the-boxplot
draft: true
categories: []
tags:
  - ggplot2
  - visualization
---

While boxplots may not be everyone's favorite way to visualize a data distribution, they are a really useful data exploration tool, and being able to customize them to your every need is pretty great. 

violin plto cartoon

`ggplot2` is great for easy out-of-the-box visualizations, and lets you do a lot of customizations once you get familiar with the grammar of it. But sometimes you want to trick out your plots with something even extra.  This post will go through some of the basics of boxplots in `ggplot` and work up to some extra ways to customize it all.

# Basic Boxplot

I'm going to use the `diamonds` dataset that comes with `ggplot2` because it has a lot of categorical and continuous variables to play around with.

```{r}
library(ggplot2)
head(diamonds)
```

First, we'll make a basic boxplot looking at price across cut.

For a standard boxplot, you set your `x` variable to the categorical variable you are interested in, and the `y` to the continuous variable. Then it is as simple as adding `geom_boxplot` to your ggplot object.

```{r}
ggplot(data=diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```

The spread in price makes this not a very useful boxplot, so I'll change the y-axis to be on the log scale using `scale_y_log10`. I also prefer to use the `theme_bw` ggplot theme.

```{r}
ggplot(data=diamonds, aes(x = cut, y = price)) +
  geom_boxplot() +
  scale_y_log10() +
  theme_bw()
```

I can add color to the boxplot to distinguish across cuts by adding an `aes` to the boxplot call. Like the hollow symbols used in `geom_point`, color refers to the outline of the boxplot, while fill will actually color the box in.

```{r}
ggplot(data=diamonds, aes(x = cut, y = price, fill = cut)) +
  geom_boxplot() +
  scale_y_log10() +
  theme_bw()
```

If you want to visualize two variables at once, you simply add the second as another aesthetic. In this case, I add clarity as an alpha, and remove the color aesthetic that was set for cut.

```{r}
ggplot(data=diamonds, aes(x = cut, y = price, alpha = clarity)) +
  geom_boxplot(fill="navyblue") +
  scale_y_log10() +
  theme_bw()
```

The thing I like about ggplot is how easy it is to add yet another marginal variable. Here, I can add the color of the diamond too.

```{r}
ggplot(data=diamonds, aes(x = cut, y = price, alpha = clarity, fill = color)) +
  geom_boxplot() +
  scale_y_log10() +
  theme_bw()
```

This now becomes basically impossible for someone to pick out, unless I make the plot very very wide. Instead, I will use another way to visualize a variable, that isn't reliant on aesthetic, `facet_wrap`. This is similar to `lattice`, and will create a seperate plot for each level of the categorical variable I choose. It won't work, however, for a continuous variable.

```{r}
ggplot(data=diamonds, aes(x = cut, y = price, alpha = clarity)) +
  geom_boxplot(fill = "navyblue") +
  scale_y_log10() +
  theme_bw() +
  facet_wrap(~color, dir = 'v')
```



Basic
Color and Fill
Dodge
Error Bars
Adding Points Behind

# Getting Fancy

# Super Fancy

Setting ranges